---
title: "Iris Dataset Statistics Lab"
author: "Alex Hurtado Desaulniers"
output: html_notebook
---

# Libraries
```{r}
library(tidyverse)
```

# Load the iris dataset


```{r}
data(iris)
```

```{r} 
# Changing names of each column
colnames(iris) <- c("SepalLen", "SepalWid", "PetalLen", "PetalWid", "Species")
```

```{r} 
# checking  first few rows
head(iris)
```

## Questions:
 1. How many numeric columns are in the dataset?
 2. How many species are included?
 3. What is the range of SepalLen? (min and max)

# Summary Statistics
Calculate mean, median, variance, SD for numeric columns

```{r}
sapply(iris[,1:4], mean)
sapply(iris[,1:4], median)
sapply(iris[,1:4], var)
sapply(iris[,1:4], sd)
```

## Questions:
1. Which variable has the largest spread?
2. Which has the smallest spread?
3. How does the mean compare to the median for PetalLen?
4. What does a large variance indicate?

# Summary by Species

```{r}
aggregate(SepalLen ~ Species, data=iris, FUN=mean)
aggregate(PetalWid ~ Species, data=iris, FUN=var)
```

## Questions:
 1. Which species has the largest SepalLen on average?
 2. Which species shows the most variability in PetalWid?
 3. What does this tell you about the characteristics of different species?

# Visualizations

### Histogram of SepalLen

```{r}
hist(iris$SepalLen,
     main="Sepal Length Distribution",
     xlab="Sepal Length (cm)",
     col="pink",
     border="black")
```

## Questions:
 1. What is the shape of the distribution (symmetric, skewed)?
 2. Are there any unusually low or high values?

# Boxplot of PetalLen by Species

```{r}
boxplot(PetalLen ~ Species,
        data=iris,
        main="Petal Length by Species",
        ylab="Petal Length (cm)",
        col=c("pink","skyblue","purple"))
```

## Questions:
 1. Which species has the longest petals?
 2. Are there any outliers visible?
 3. How do the distributions differ between species?

# Correlation Matrix

```{r}
cor_matrix <- cor(iris[,1:4])
cor_matrix
```

## Questions:
 1. Which two variables are most strongly correlated?
 2. Are there any weak or negative correlations?
 3. How might these correlations help distinguish species?

# Interpretation 

Write a short paragraph (3-5 sentences) answering:

- Which variables best distinguish species?
- How could a scientist use these statistics in real life?
- Do the histograms, boxplots, and correlations tell a consistent story?


